---
title: "PML Project"
author: "Arunkumar"
date: "15/04/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
##Practical Machine Learning - Week 4 Assignment 
  

### Background  (reproduced from assignment page)


Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).

Data

The training data for this project are available here: https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv

The test data are available here:https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv

The data for this project come from this source: http://web.archive.org/web/20161224072740/http:/groupware.les.inf.puc-rio.br/har.

#### Assignment Requirements 
The goal of this project is to predict the manner in which they did the exercise. This is the "classe" variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.  
1. Your submission should consist of a link to a Github repo with your R markdown and compiled HTML file describing your analysis. Please constrain the text of the writeup to < 2000 words and the number of figures to be less than 5. It will make it easier for the graders if you submit a repo with a gh-pages branch so the HTML page can be viewed online (and you always want to make it easy on graders.  
2. You should also apply your machine learning algorithm to the 20 test cases available in the test data above. Please submit your predictions in appropriate format to the programming assignment for automated grading. See the programming assignment for additional details.  

#### Loading required libraries  
  
```{r warning=FALSE, error=FALSE}
library(rattle)
library(caret)
library(rpart)
library(rpart.plot)
library(corrplot)
library(randomForest)
library(RColorBrewer)
```  
Finally, load the same seed with the following line of code:  
```{r warning=FALSE, error=FALSE}
set.seed(56789)
```  

#### Getting Data  
Setting the working directory, download data and read files
```{r warning=FALSE, error=FALSE}
setwd("D:/Arun/Coursera/PML")

trainUrl <-"https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"
trainFile <- "./data/pml-training.csv"
testFile  <- "./data/pml-testing.csv"
if (!file.exists("./data")) {
  dir.create("./data")
}
if (!file.exists(trainFile)) {
  download.file(trainUrl, destfile = trainFile, method = "curl")
}
if (!file.exists(testFile)) {
  download.file(testUrl, destfile = testFile, method = "curl")
}
rm(trainUrl)
rm(testUrl)
```  

#### Reading Data  
Read the files to data frames.  
```{r warning=FALSE, error=FALSE}
trainRaw <- read.csv(trainFile)
testRaw <- read.csv(testFile)
dim(trainRaw)
dim(testRaw)
rm(trainFile)
rm(testFile)
```  
The training data set contains `r dim(trainRaw)[1]` observations and `r dim(trainRaw)[2]` variables, while the testing data set contains `r dim(testRaw)[1]` observations and `r dim(testRaw)[2]` variables. The `classe` variable in the training set is the outcome to predict.  

#### Cleaning Data  
 
Cleanse data to remove observations with missing values and other unwanted variables

1. Clean the <b>Near Zero Variance</b> Variables.  
```{r warning=FALSE, error=FALSE}
NZV <- nearZeroVar(trainRaw, saveMetrics = TRUE)
head(NZV, 20)
training01 <- trainRaw[, !NZV$nzv]
testing01 <- testRaw[, !NZV$nzv]
dim(training01)
dim(testing01)
rm(trainRaw)
rm(testRaw)
rm(NZV)
```  

2. Removing few columns of the dataset that do not add value to the accelerometer measurements.  
```{r warning=FALSE, error=FALSE}
regex <- grepl("^X|timestamp|user_name", names(training01))
training <- training01[, !regex]
testing <- testing01[, !regex]
rm(regex)
rm(training01)
rm(testing01)
dim(training)
dim(testing)
```  

3. Removing `NA's`.  
```{r warning=FALSE, error=FALSE}
cond <- (colSums(is.na(training)) == 0)
training <- training[, cond]
testing <- testing[, cond]
rm(cond)
```  

The cleaned training dataset contains `r dim(training)[1]` observations and `r dim(training)[2]` variables, and the testing dataset contains `r dim(testing)[1]` observations and `r dim(testing)[2]` variables.  

Correlation Matrix of columns in the training dataset.  
```{r warning=FALSE, error=FALSE}
corrplot(cor(training[, -length(names(training))]), method = "color", tl.cex = 0.5)
```  

#### Partitioning Training Set  
We split the cleaned training dataset into a pure training data set (70%) and a validation data set (30%). Validation dataset to be used for cross validation later.
```{r warning=FALSE, error=FALSE}
set.seed(56789) # For reproducibile purpose
inTrain <- createDataPartition(training$classe, p = 0.70, list = FALSE)
validation <- training[-inTrain, ]
training <- training[inTrain, ]
rm(inTrain)
```  
The Dataset now consists of `r dim(training)[2]` variables with the observations divided as following:  
1. Training Data: `r dim(training)[1]` observations.  
2. Validation Data: `r dim(validation)[1]` observations.  
3. Testing Data: `r dim(testing)[1]` observations.  

#### Data Modelling  

#### Decision Tree  
Fitting a predictive model for activity recognition using <b>Decision Tree</b> algorithm.  
```{r warning=FALSE, error=FALSE}
modelTree <- rpart(classe ~ ., data = training, method = "class")
prp(modelTree)
```  

Estimating the model's performance for <b>validation</b> data set.  
```{r warning=FALSE, error=FALSE}
predictTree <- predict(modelTree, validation, type = "class")
confusionMatrix(validation$classe, predictTree)
accuracy <- postResample(predictTree, validation$classe)
ose <- 1 - as.numeric(confusionMatrix(validation$classe, predictTree)$overall[1])
rm(predictTree)
rm(modelTree)
```  

The estimated accuracy of Random Forest Model is `r accuracy[1]*100`% and the estimated Out-of-Sample Error is `r ose*100`%.  

#### Random Forest
We fit a predictive model for activity recognition using <b>Random Forest</b> algorithm with <b>5-fold cross validation</b>   
  
```{r warning=FALSE, error=FALSE}
modelRF <- train(classe ~ ., data = training, method = "rf", trControl = trainControl(method = "cv", 5), ntree = 250)
modelRF
```  

Estimating the model's performance for <b>validation</b> data set.  
```{r warning=FALSE, error=FALSE}
predictRF <- predict(modelRF, validation)
confusionMatrix(validation$classe, predictRF)
accuracy <- postResample(predictRF, validation$classe)
ose <- 1 - as.numeric(confusionMatrix(validation$classe, predictRF)$overall[1])
rm(predictRF)
```  

The estimated accuracy of the Random Forest Model is `r accuracy[1]*100`% and the estimated out-of-sample error is `r ose*100`%.  
Random Forests yielded better Results, as expected!  

#### Predicting the Manner of exercise for test dataset  
Now, we apply the <b>Random Forest</b> model to the original testing dataset downloaded from the data source. We remove the problem_id column first.  
```{r warning=FALSE, error=FALSE}
rm(accuracy)
rm(ose)
predict(modelRF, testing[, -length(names(testing))])
```  

#### Generate answer files for the assignment quiz 
Function to generate files with predictions for quiz.  
```{r warning=FALSE, error=FALSE}
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("./problem_id_",i,".txt")
    write.table(x[i], file = filename, quote = FALSE, row.names = FALSE, col.names = FALSE)
  }
}
```  
writing the files.  
```{r warning=FALSE, error=FALSE}
pml_write_files(predict(modelRF, testing[, -length(names(testing))]))
rm(modelRF)
rm(training)
rm(testing)
rm(validation)
rm(pml_write_files)
```  